On hyperparameter tuning through Lipschitz global optimization

### 2024-03-17 2024/ITMO paper/sn-article.tex
resubmission of ITMO paper in 2024

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%%\usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

%% Русский язык - закомментировать в финальной версии
\usepackage[russian]{babel}

%% advanced mathematics
\usepackage{amsmath}
\usepackage[ruled]{algorithm2e}
\usepackage{multirow}

\usepackage{graphicx}
\usepackage[table,xcdraw]{xcolor}
\usepackage{amssymb}

\usepackage{hyperref}
\usepackage[normalem]{ulem}

\journal{Knowledge-Based Systems}

@@

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%% \fntext[label3]{}

\title{On hyperparameter tuning through Lipschitz global optimization} % and dimensionality reduction}?
%% О настройке гиперпараметров с помощью липшицевой глобальной оптимизации

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\author[UNN,ITMO]{Konstantin Barkalov\corref{cor}}
\cortext[cor]{Corresponding author: \url{konstantin.barkalov@itmm.unn.ru}}
\author[UNN,ITMO]{Denis Karchkov}
\author[UNN,ITMO]{Evgeny Kozinov}
\author[UNN,ITMO]{Ilya Lebedev}
\author[UNN,ITMO]{Denis Rodionov}
\author[UNN,ITMO]{Marina Usova}

\affiliation[UNN]{organization={Lobachevsky University},%Department and Organization
            addressline={Gagarin Ave. 23}, 
            city={Nizhni Novgorod},
            postcode={603022}, 
            %state={},
            country={Russia}}

\affiliation[ITMO]{organization={ITMO University},%Department and Organization
            addressline={Lomonosova St. 9}, 
            city={St. Petersburg},
            postcode={191002}, 
            %state={},
            country={Russia}}

\begin{abstract}
%% Text of abstract

The quality of machine learning methods is substantially affected by their hyperparameters, while the evaluation of the quality criterion is a time-consuming operation. Therefore, it is important to develop intelligent methods for selecting optimal values of hyperparameters that require a small number of search trials. In this paper, we propose a new approach to hyperparameter tuning based on the ideas of Lipschitz global optimization. In the framework of this approach, the solution of problems with several parameters is reduced to solving equivalent one-dimensional problems. The reduction is based on the use of space-filling curves (Peano curves). These approaches are implemented in the open-source framework of \textit{i}ntelligent \textit{opt}imization methods \textcolor[rgb]{1,0,0}{(iOpt)}.  To demonstrate the advantages of iOpt, we compare it with the well-known Optuna and HyperOpt frameworks when tuning hyperparameters of various machine learning methods on \textcolor[rgb]{1,0,0}{\sout{a} the}  representative \textcolor[rgb]{1,0,0}{ \sout{set of} } datasets. The results show that Lipschitz global optimization methods provide comparable (in terms of quality) results in a significantly shorter time compared to known hyperparameter tuning algorithms.

\end{abstract}

%%Graphical abstract
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}

%%Research highlights
%\begin{highlights}
%\item A framework of Lipschitz optimization methods is developed for the tuning of hyperparameters
%\item Dimensionality reduction based on Peano space-filling curves is used to solve multidimensional problems
%\item A scheme is proposed for reducing the problems, some of whose parameters are categorical, to a single problem with continuous parameters
%\item The developed methods (unlike the known methods of hyperparameter tuning) are deterministic, which ensures stable results at different runs of the algorithms 
%
%\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Lipschitz optimization \sep Global optimization \sep Multiextremal functions \sep Hyperparameter tuning

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%\linenumbers

%% main text
\section{Introduction}
\label{sec_intro}

Currently, complex scientific and/or industrial problems are most often solved by numerical methods. Most of such methods, for example, machine learning methods, have hyperparameters, and the choice of specific values of these hyperparameters can determine the efficiency (in a given metric) of the obtained solution under certain constraints on the computation time \cite{Hutter2019,nikitin2021}. The fact that in many cases the difference in the efficiency of the solution obtained with different values of hyperparameters can be quite significant, has resulted in the emergence of the hyperparameter optimization (HPO) problem, as well as the research of approaches to its solution that are implemented in the frameworks for automatic selection of hyperparameters \cite{Tune, optuna, hyperopt,Sherpa}. From the mathematical point of view, the HPO problem corresponds to a global optimization problem with fixed bounds on the variation of hyperparameters.  In certain cases, some of the hyperparameters may be categorical, i.e., they can only assume values from a certain, usually small, number of variants, which necessitates the use of optimization methods capable of solving problems with discrete parameters.

@@

\textcolor[rgb]{1,0,0}{Отличительной особенностью фреймворка iOpt является использование for hyperparameters tuning методов липшицевой оптимизации, которые являются детерминированными, обладают быстрой сходимостью, а также возможностью адаптации к различного рода специальным постановкам задач.
Например, рассматриваемые методы допускают обобщение для работы с зашумленными функциями \cite{Sergeyev2020}. В данной работе они были адаптированы для решения HPO problems, в которых часть параметров часто являются дискретными и/или категориальными.}

@@

\section{Related work}
\label{sec_rel}

@@

The simplest way to find the global optimum is the exhaustive search method (search on a uniform \cite{Bao2006} or random \cite{Bergstra2012} grid). 
By selecting the number of analyzed variants for each of the hyperparameters, searching all the resulting combinations and choosing the best of them in a given metric, we obtain an estimate of the optimum with a certain accuracy \cite{Nevendra2022}. The exhaustive search method is easy to implement, it has perfect parallelism, but it has a significant drawback (an exponential growth of the number of grid nodes as the number of hyperparameters increases). Thus, exhaustive search can be used as a method for solving HPO problems only in ``simple'' problems where both the number of hyperparameters (one or two) and the time for solving the initial problem with each selected combination of hyperparameters are small.

Metaheuristic (genetic and similar) algorithms are an alternative to the exhaustive search when the number of hyperparameters increases \cite{Opara2019}. These algorithms are widely used in the absence of a formula-based description of the model under study (``black-box'' model), which is typical for the considered HPO problems \cite{Zhou2021,Yang2022}. Metaheuristic algorithms only weakly depend on the number of hyperparameters, they use information from previous iterations to perform the current one, but due to the randomness inherent in the methods, they guarantee the finding of the global optimum only in the probabilistic sense.

Another approach to finding the global optimum is Bayesian optimization, which is also used to solve ``black-box'' problems \cite{Frazier2018,Archetti2019}. In fact, Bayesian optimization methods build a stochastic model of the function being optimized. The model is iteratively updated on the basis of the information accumulated in the process of searching for the optimum, making it possible to estimate the most probable position of the global optimum at each iteration. Bayesian optimization methods have significantly higher efficiency compared to the exhaustive search, but they are also affected by the ``curse of dimensionality'', albeit to a lesser extent.

In well-known hyperparameter tuning frameworks, the Tree-Structured Parzen Estimator (TPE) algorithm is widely used \cite{hyperopt,NIPS2011}. TPE is an iterative algorithm, conceptually close to Bayesian optimization methods. Before the algorithm starts, a search region in the hyperparameter space must be specified and the function to be optimized must be defined. In the course of its work, the method performs trials in the search region and then uses these trials to construct estimates of the distribution of the best and all other results in the hyperparameter space. Next, the combination of hyperparameters that gives the maximum expected improvement is searched for. These hyperparameters are then evaluated on the objective function. The described process is repeated until a specified number of iterations is attained or until the time constraint is reached.

The good results achieved by Bayesian optimization \cite{Joy2020} and TPE \cite{Watanabe2022a,Watanabe2022b} methods are based on the a priori assumption that the function corresponds to a certain stochastic model, such as a Gaussian process \cite{Rasmussen2005}. However, there are other assumptions about the shape of the function in global optimization that yield remarkable results.

One of such assumptions about the problem to be solved is the assumption of boundedness of relative changes in the objective function. In this case, the function is said to satisfy the Lipschitz condition, and the problem to be solved is called a Lipschitz global optimization problem. For the class of Lipschitz global optimization problems, a number of efficient algorithms have been developed \cite{Jones2021,Paulavicius2020,Strongin2020,Sergeyev2017,PaulaviciusZilinskas2014}, which surpass many other global optimization methods \cite{Sergeyev2018} and have shown their applicability in a number of applications \cite{Kvasov2008,CANDELIERI2019,Gubaydullin2021}.

The iOpt framework contains an implementation of Lipschitz optimization algorithms, which are an extension of the information-statistical global search algorithm described in detail in \cite{Strongin2000,Sergeyev2013}. In particular, the methods have been adapted to solve problems with partially integer parameters, which are very common in HPO problems. The experimental results presented in Section \ref{sec_exp} show that the iOpt framework is not inferior in HPO problems to well-known frameworks having the same purpose (such as Optuna \cite{optuna} and HyperOpt \cite{hyperopt}).

\section{Lipschitz optimization} 
\label{sec_lip}

@@

We will also assume that the objective function $\varphi(y)$ satisfies the Lipschitz condition in the search domain $D$, i.e.
\begin{equation} \label{f_lip}
	\left| \varphi(y')-\varphi(y'') \right| \leq L\left\| y' - y''  \right\| , \; y',y'' \in D, \; 0<L<\infty.
\end{equation}
Here $ \left\| \cdot \right\|$ denotes the Euclidean norm, and the constant $L$ is not known a priori and is subject to estimation in the process of solution.

The function $\varphi(y)$ is assumed to be multiextremal and specified in the form of a ``black box'', i.e., of some algorithm for computing its values. Moreover, it is assumed that each search trial (i.e., computing the value of the function at a point in the admissible region) requires significant computational resources. This problem formulation fully corresponds to the task of tuning the hyperparameters of machine learning methods \cite{Joy2020,Wang2021}.

@@

  By virtue of these inequalities, the values of the function at the points of the interval $[x', x'']$ must lie inside the region bounded by the straight lines passing through the points $(x', f(x'))$ and $(x'', f(x''))$ with slopes $L$ and $-L$ (see Fig. \ref{fig1}).

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{Fig1.pdf}
\caption{The values of the function $f(x)$ must be inside the area highlighted by colour.} \label{fig1}
\end{figure}

@@

\[
F(x) = \max\left\{f(x') - L(x-x'),f(x') + L(x-x'')\right\}, \; x\in [x', x''].
\]

@@

Lipschitz global optimization methods use this idea in their computational rules; the proof of their convergence is also based on this idea (see, e.g.,
\cite{Jones2021,PaulaviciusZilinskas2014,Sergeyev2013,Evtushenko2013}).

@@

A well-known approach to solving the multidimensional problem (\ref{f_func}) is to reduce it to a one-dimensional problem and then apply efficient one-dimensional global optimization algorithms (see \cite{Strongin2000,Sergeyev2013}). A well-proven approach here is to reduce the dimensionality of the problem by means of a Peano-Hilbert curve $y(x), x \in [0, 1]$.  Such a curve uniquely and continuously maps the interval $[0, 1]$ onto the hyperinterval $D$ of (\ref{f_D}), i.e., it fills the entire region $D$.

@@

It is known (see \cite{Strongin2000}) that if a multidimensional function  $\varphi(y), \; y \in D$,  satisfies the Lipschitz condition with the constant $L$, then the reduced one-dimensional function $\varphi(y(x)), \; x \in [0,1]$ satisfies the H{\"o}lder condition
\begin{equation}\label{holder}
\left|\varphi(y(x'))-\varphi(y(x''))\right|\leq H\left|x'-x''\right|^{1/N}, \; x',x''\in[0,1].
\end{equation}
Here, $N$ is the dimensionality of the original problem, and the coefficient
$ H=2 L \sqrt{N+3}$.

Remark 1.  The Lipshitz condition will not be satisfied for the reduced one-dimensional function $\varphi(y(x))$. However, many one-dimensional Lipschitz optimization algorithms can be generalized to the case of H{\"o}lder function minimization. A concrete example of such a generalization is given in subsection \ref{sec_GSA}.

\begin{figure}
\center
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig2a.jpg} }
\end{minipage}
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig2b.jpg} }
\end{minipage}
\caption{Evolvents $y_m(x)$ with $m=4$ for dimensions $N=2$ and $N=3$.}\label{fig:Peano}
\end{figure}

Remark 2. The theoretical Peano-Hilbert curve $y(x)$ is defined as a limiting object. Numerical algorithms use evolvents $y_m(x)$, that approximate the true curve with a given density level $2^{-m}$, which depends on the required search accuracy. Efficient schemes for computing several types of evolvents are described in detail in \cite{Sergeyev2013}. For illustration, Fig.~\ref{fig:Peano} shows graphs corresponding to the evolvents $y_m(x)$ with density $m=4$ for dimensions $N=2$ and $N=3$.

\subsection{The problem of discrete parameters}
\label{sec_discr}

Of particular interest are problems where some of the parameters can assume values from a predetermined finite set, since it is more difficult to construct estimates of the optimum for these problems than for problems with continuous parameters. Many publications address the methods for solving mixed-parameter problems (see, for example, reviews \cite{Burer2012,Boukouvala2016}). Known deterministic methods for solving problems of this class are centered on solving linear or convex problems \cite{Lee2012}. It is suggested that complex multiextremal problems should be solved using heuristic approaches \cite{Belotti2013}.

@@

@@

\section{iOpt framework} 
\label{sec_iOpt}

@@

\begin{figure}[t]
\centering
\includegraphics[width=1.0\textwidth]{scheme.pdf}
\caption{iOpt general structure.} \label{fig_iOpt}
\end{figure}

@@

\begin{table}

@@

\end{table}

>>>>

%Version 3 October 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%======================================================%%
%% to compile with pdflatex/xelatex use pdflatex option %%
%%======================================================%%

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove Numbered in the optional parenthesis. 
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
\documentclass[iicol]{sn-jnl}
%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

%%%% Standard Packages

%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%% Русский язык - закомментировать в финальной версии
\usepackage[russian]{babel}
%% Зачеркивание - закомментировать в финальной версии
\usepackage[normalem]{ulem}

\usepackage[round]{natbib}
\bibliographystyle{abbrvnat}
%\setcitestyle{authoryear,open={(},close={)}} %Citation-related commands

%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

@@

\title[Article Title]{On hyperparameter tuning through Lipschitz global optimization}
%% О настройке гиперпараметров с помощью липшицевой глобальной оптимизации

%%=============================================================%%
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
%%  \sfx{IV}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2]{\fnm{Konstantin} \sur{Barkalov}}\email{konstantin.barkalov@itmm.unn.ru}

\author[1,2]{\fnm{Denis} \sur{Karchkov}}

\author[1,2]{\fnm{Evgeny} \sur{Kozinov}}

\author[1,2]{\fnm{Ilya} \sur{Lebedev}}

%\author[1,2]{\fnm{Denis} \sur{Rodionov}}

\author[1,2]{\fnm{Marina} \sur{Usova}}

\affil[1]{
%\orgdiv{Department}, 
\orgname{Lobachevsky University}, \orgaddress{\street{Gagarin Ave. 23}, \city{Nizhni Novgorod}, \postcode{603022},
%\state{State},
\country{Russia}}}

\affil[2]{
%\orgdiv{Department},
\orgname{ITMO University}, \orgaddress{\street{Lomonosova St. 9}, \city{St. Petersburg}, \postcode{191002}, 
%\state{State},
\country{Russia}}}

%%==================================%%
%% Sample for unstructured abstract %%
%%==================================%%

\abstract{The quality of machine learning methods is substantially affected by their hyperparameters, while the evaluation of the quality criterion is a time-consuming operation. Therefore, it is important to develop intelligent methods for selecting optimal values of hyperparameters that require a small number of search trials. In this paper, we propose a new approach to hyperparameter tuning based on the ideas of Lipschitz global optimization. In the framework of this approach, the solution of problems with several parameters is reduced to solving equivalent one-dimensional problems. The reduction is based on the use of space-filling curves (Peano curves). These approaches are implemented in the open-source framework of \textit{i}ntelligent \textit{opt}imization methods \textcolor[rgb]{1,0,0}{(iOpt)}.  To demonstrate the advantages of iOpt, we compare it with the well-known Optuna and HyperOpt frameworks when tuning hyperparameters of various machine learning methods on \textcolor[rgb]{1,0,0}{\sout{a} the}  representative \textcolor[rgb]{1,0,0}{ \sout{set of} } datasets. The results show that Lipschitz global optimization methods provide comparable (in terms of quality) results in a significantly shorter time compared to known hyperparameter tuning algorithms.}

%%================================%%
%% Sample for structured abstract %%
%%================================%%

% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}

\keywords{Lipschitz optimization, Global optimization, Multiextremal functions, Hyperparameter tuning}

%%\pacs[JEL Classification]{D8, H51}

%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}

\maketitle
\section{Introduction}\label{sec_intro}

Currently, complex scientific and/or industrial problems are most often solved by numerical methods. Most of such methods, for example, machine learning methods, have hyperparameters, and the choice of specific values of these hyperparameters can determine the efficiency (in a given metric) of the obtained solution under certain constraints on the computation time \citep{Hutter2019,nikitin2021}. The fact that in many cases the difference in the efficiency of the solution obtained with different values of hyperparameters can be quite significant, has resulted in the emergence of the hyperparameter optimization (HPO) problem, as well as the research of approaches to its solution that are implemented in the frameworks for automatic selection of hyperparameters \citep{Tune,optuna,hyperopt,Sherpa}. From the mathematical point of view, the HPO problem corresponds to a global optimization problem with fixed bounds on the variation of hyperparameters.  In certain cases, some of the hyperparameters may be categorical, i.e., they can only assume values from a certain, usually small, number of variants, which necessitates the use of optimization methods capable of solving problems with discrete parameters.

@@

\textcolor[rgb]{1,0,0}{Отличительной особенностью фреймворка iOpt является использование for hyperparameters tuning методов липшицевой оптимизации, которые являются детерминированными, обладают быстрой сходимостью, а также возможностью адаптации к различного рода специальным постановкам задач.
Например, рассматриваемые методы допускают обобщение для работы с зашумленными функциями \citep{Sergeyev2020}. В данной работе они были адаптированы для решения HPO problems, в которых часть параметров часто являются дискретными и/или категориальными.}

@@

\section{Related work}\label{sec_rel}

@@

The simplest way to find the global optimum is the exhaustive search method (search on a uniform \citep{Bao2006} or random \citep{Bergstra2012} grid). 
By selecting the number of analyzed variants for each of the hyperparameters, searching all the resulting combinations and choosing the best of them in a given metric, we obtain an estimate of the optimum with a certain accuracy \citep{Nevendra2022}. The exhaustive search method is easy to implement, it has perfect parallelism, but it has a significant drawback (an exponential growth of the number of grid nodes as the number of hyperparameters increases). Thus, exhaustive search can be used as a method for solving HPO problems only in ``simple'' problems where both the number of hyperparameters (one or two) and the time for solving the initial problem with each selected combination of hyperparameters are small.

Metaheuristic (genetic and similar) algorithms are an alternative to the exhaustive search when the number of hyperparameters increases \citep{Opara2019}. These algorithms are widely used in the absence of a formula-based description of the model under study (``black-box'' model), which is typical for the considered HPO problems \citep{Zhou2021,Yang2022}. Metaheuristic algorithms only weakly depend on the number of hyperparameters, they use information from previous iterations to perform the current one, but due to the randomness inherent in the methods, they guarantee the finding of the global optimum only in the probabilistic sense.

Another approach to finding the global optimum is Bayesian optimization, which is also used to solve ``black-box'' problems \citep{Frazier2018,Archetti2019}. In fact, Bayesian optimization methods build a stochastic model of the function being optimized. The model is iteratively updated on the basis of the information accumulated in the process of searching for the optimum, making it possible to estimate the most probable position of the global optimum at each iteration. Bayesian optimization methods have significantly higher efficiency compared to the exhaustive search, but they are also affected by the ``curse of dimensionality'', albeit to a lesser extent.

In well-known hyperparameter tuning frameworks, the Tree-Structured Parzen Estimator (TPE) algorithm is widely used \citep{hyperopt,NIPS2011}. TPE is an iterative algorithm, conceptually close to Bayesian optimization methods. Before the algorithm starts, a search region in the hyperparameter space must be specified and the function to be optimized must be defined. In the course of its work, the method performs trials in the search region and then uses these trials to construct estimates of the distribution of the best and all other results in the hyperparameter space. Next, the combination of hyperparameters that gives the maximum expected improvement is searched for. These hyperparameters are then evaluated on the objective function. The described process is repeated until a specified number of iterations is attained or until the time constraint is reached.

The good results achieved by Bayesian optimization \citep{Joy2020} and TPE \citep{Watanabe2022a,Watanabe2022b} methods are based on the a priori assumption that the function corresponds to a certain stochastic model, such as a Gaussian process \citep{Rasmussen2005}. However, there are other assumptions about the shape of the function in global optimization that yield remarkable results.

One of such assumptions about the problem to be solved is the assumption of boundedness of relative changes in the objective function. In this case, the function is said to satisfy the Lipschitz condition, and the problem to be solved is called a Lipschitz global optimization problem. For the class of Lipschitz global optimization problems, a number of efficient algorithms have been developed \citep{Jones2021,Paulavicius2020,Strongin2020,Sergeyev2017,PaulaviciusZilinskas2014}, which surpass many other global optimization methods \citep{Sergeyev2018} and have shown their applicability in a number of applications \citep{Kvasov2008,CANDELIERI2019,Gubaydullin2021}.

The iOpt framework contains an implementation of Lipschitz optimization algorithms, which are an extension of the information-statistical global search algorithm described in detail in \citet{Strongin2000,Sergeyev2013}. In particular, the methods have been adapted to solve problems with partially integer parameters, which are very common in HPO problems. The experimental results presented in Section \ref{sec_exp} show that the iOpt framework is not inferior in HPO problems to well-known frameworks having the same purpose (such as Optuna \citep{optuna} and HyperOpt \citep{hyperopt}).

\section{Lipschitz optimization}\label{sec_lip}

@@

We will also assume that the objective function $\varphi(y)$ satisfies the Lipschitz condition in the search domain $D$, i.e.
\begin{equation} \label{f_lip}
	\left| \varphi(y')-\varphi(y'') \right| \leq L\left\| y' - y''  \right\| , \,
\end{equation}
where $y',y'' \in D, \; 0<L<\infty$, and $ \left\| \cdot \right\|$ is denotes the Euclidean norm, and the constant $L$ is not known a priori and is subject to estimation in the process of solution.

The function $\varphi(y)$ is assumed to be multiextremal and specified in the form of a ``black box'', i.e., of some algorithm for computing its values. Moreover, it is assumed that each search trial (i.e., computing the value of the function at a point in the admissible region) requires significant computational resources. This problem formulation fully corresponds to the task of tuning the hyperparameters of machine learning methods \citep{Joy2020,Wang2021}.

@@

  By virtue of these inequalities, the values of the function at the points of the interval $[x', x'']$ must lie inside the region bounded by the straight lines passing through the points $(x', f(x'))$ and $(x'', f(x''))$ with slopes $L$ and $-L$ (see Fig.~\ref{fig1}).

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{Fig1.pdf}
\caption{The values of the function $f(x)$ must be inside the area highlighted by colour.} \label{fig1}
\end{figure}

@@

\[
F(x) = \max\left\{f(x') - L(x-x'),f(x') + L(x-x'')\right\},
\] 
where $\; x\in [x', x'']$.

@@

Lipschitz global optimization methods use this idea in their computational rules; the proof of their convergence is also based on this idea (see, e.g.,
\citet{Jones2021,PaulaviciusZilinskas2014,Sergeyev2013,Evtushenko2013}).

@@

A well-known approach to solving the multidimensional problem (\ref{f_func}) is to reduce it to a one-dimensional problem and then apply efficient one-dimensional global optimization algorithms (see \citet{Strongin2000,Sergeyev2013}). A well-proven approach here is to reduce the dimensionality of the problem by means of a Peano-Hilbert curve $y(x), x \in [0, 1]$.  Such a curve uniquely and continuously maps the interval $[0, 1]$ onto the hyperinterval $D$ of (\ref{f_D}), i.e., it fills the entire region $D$.

@@

It is known (see \citet{Strongin2000}) that if a multidimensional function  $\varphi(y), \; y \in D$,  satisfies the Lipschitz condition with the constant $L$, then the reduced one-dimensional function $\varphi(y(x)), \; x \in [0,1]$ satisfies the H{\"o}lder condition
\begin{equation}\label{holder}
\left|\varphi(y(x'))-\varphi(y(x''))\right|\leq H\left|x'-x''\right|^{1/N}.
\end{equation}
Here, $x',x''\in[0,1]$, $N$ is the dimensionality of the original problem, and the coefficient
$ H=2 L \sqrt{N+3}$.

\begin{remark}
The Lipshitz condition will not be satisfied for the reduced one-dimensional function $\varphi(y(x))$. However, many one-dimensional Lipschitz optimization algorithms can be generalized to the case of H{\"o}lder function minimization. A concrete example of such a generalization is given in subsection \ref{sec_GSA}.

\begin{figure}
\center
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig2a.jpg} }
\end{minipage}
\begin{minipage}{0.45\linewidth}
\center{\includegraphics[width=0.9\linewidth]{fig2b.jpg} }
\end{minipage}
\caption{Evolvents $y_m(x)$ with $m=4$ for dimensions $N=2$ and $N=3$.}\label{fig:Peano}
\end{figure}   
\end{remark}

\begin{remark}
The theoretical Peano-Hilbert curve $y(x)$ is defined as a limiting object. Numerical algorithms use evolvents $y_m(x)$, that approximate the true curve with a given density level $2^{-m}$, which depends on the required search accuracy. Efficient schemes for computing several types of evolvents are described in detail in \citet{Sergeyev2013}. For illustration, Fig.~\ref{fig:Peano} shows graphs corresponding to the evolvents $y_m(x)$ with density $m=4$ for dimensions $N=2$ and $N=3$.
\end{remark}

\subsection{The problem of discrete parameters}\label{sec_discr}

Of particular interest are problems where some of the parameters can assume values from a predetermined finite set, since it is more difficult to construct estimates of the optimum for these problems than for problems with continuous parameters. Many publications address the methods for solving mixed-parameter problems (see, for example, reviews \citet{Burer2012,Boukouvala2016}). Known deterministic methods for solving problems of this class are centered on solving linear or convex problems \citep{Lee2012}. It is suggested that complex multiextremal problems should be solved using heuristic approaches \citep{Belotti2013}.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=1.0\textwidth]{scheme.pdf}
\caption{iOpt general structure.} \label{fig_iOpt}
\end{figure*}

@@

\section{iOpt framework}\label{sec_iOpt}

@@

@@

\begin{table*}

@@

\end{table*}
